A Deep Learning Approach for Gait Event Detection from a Single Shank-Worn IMU: Validation in Healthy and Neurological Cohorts

### 2022-04-21 paper/MDPI_template/article.tex
Updated draft with generic tcn architecture

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Many algorithms use 3D accelerometer and/or gyroscope data from inertial measurement unit (IMU) sensors to detect gait events (i.e., initial and final foot contact). However, these algorithms often require knowledge about sensor orientation and use empirically derived thresholds. As sensor-to-segment alignment cannot always be controlled for in ambulatory assessments, methods are needed that require little knowledge on sensor location and orientation, e.g., a convolutional neural network-based deep learning model. Therefore, 160 participants from healthy and neurologically diseased cohorts walked 5 meter distances at slow, preferred, and fast walking speed, while data were collected from IMUs on the left and right ankle and shank. Gait events were detected and stride parameters were extracted using a deep learning model and an optoelectronic motion capture (OMC) system for reference. Results showed a high detection rate for both initial and final contacts across sensor locations (recall $\ge 92\%$, precision $\ge 97\%$). Time agreement was excellent as witnessed from the median time error (0.005 s) and corresponding inter-quartile range (0.020 s). The extracted stride-specific parameters were in good agreement with parameters derived from the OMC system (maximum mean difference 0.003 s, and corresponding maximum limits of agreement (-0.049 s, 0.051 s) for a $95\%$ confidence level.) Thus, the deep learning approach was considered a valid approach for detecting gait events and extracting stride-specific parameters with little knowledge on exact IMU location and orientation in conditions with and without walking pathologies due to neurological diseases.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or conversion from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}. The beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or marker-based optoelectronic motion capture systems (henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is increasingly more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

Previous research suggests that gait event detection is more accurate using an IMU worn on a lower limb (e.g., shank or foot) compared to an IMU worn on the low back \cite{BenMansour2015,Storm2016,Panebianco2018}. Now, in order to get from abstract IMU sensor readings to clinically relevant gait parameters (e.g., from accelerations and angular velocities to stride times) \cite{Hannink2016}, different algorithmic approaches have been developed in the last twenty years of clinical gait research. A recent study has evaluated a cross-section of these algorithms for different sensor locations on the lower leg and foot \cite{Niswander2021}. The algorithms were categorized according to which signals were analyzed, for example the angular velocity about the medio-lateral axis, or the accelerations along vertical and antero-posterior axes. This means the sensor readings need to be linked with the anatomical axes, that is, one needs to know which sensor axis aligns with for example the medio-lateral axis. In most approaches, it is simply assumed that due to sensor attachment the sensor axis aligns roughly with the anatomical axis of interest (\cite{Salarian2004,Sabatini2005,Jasiewicz2006,Catalfamo2010,Trojaniello2014,Maqbool2016,Romijnders2021}) or an additional calibration procedure (e.g., \cite{Ferraris1995}) is required (\cite{Greene2010,Niswander2021}). In ambulatory assessments however, study participants often attach the sensor themselves, for example after showering, and therefore the sensor location and alignment cannot be controlled for. Furthermore, it is unlikely that each time the sensor is (re-)attached study participants, especially those with gait deficits, perform a calibration procedure that usually consists of holding a pre-defined pose and performing some known movement sequences \cite{Leineweber2019,Pacher2020}.

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera optoelectronic motion capture system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}.

@@

All participant data data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}). The training and validation set were used to train an optimal deep learning model. Test set data were not used for model training or hyperparameter tuning. The results on the model's performance were only based on the test set, and therefore reflect how good the model generalizes to new, \emph{unseen} data.

@@

\subsection{Model}
\subsubsection{Model Architecture}
The generic architecture for the deep learning model was based on a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of a sequence of residual blocks with exponentially increasing dilation factor \cite{YuKoltun2016,Bai2018}. Each residual block was built from two sequences of a dilated convolutional layer \cite{YuKoltun2016}, a batch normalization layer \cite{Ioffe2015}, a rectified linear unit (ReLU) activation layer, and a dropout layer \cite{Srivastava2014} (Figure). The model was built in Python \cite{VanRossum2009} using the high-level TensorFlow API Keras \cite{Remy2020,Chollet2015}.

@@

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived stride-specific gait parameters.
\subsubsection{Overall detection performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated evens were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many events that were detected, were actually not annotated (false positives, FP). From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} 
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} 
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \cdot \textrm{precision}}{\textrm{recall} + \textrm{precision}} 
	\label{eqn:f1_score}
\end{equation}

\subsubsection{Time Error}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), the time error between the annotated and detect gait event was defined as:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Stride Parameters}
For those trials for which all gait events were detected, and no false positives were detected, the stride time, stance time and swing were calculated. Stride was the time between two successive ICs of the same foot. Stance time was the time between a FC and the preceding IC of the same foot. Swing time was the time between the IC following the last FC of the same foot.

@@

\subsection{Overall detection performance}
The performance of detecting ICs and FCs was objectively quantified by the number of TPs, the number of FNs, and the number of FPs. From these numbers, recall, precision and F$_{1}$ score were calculated (Table \ref{tab:overall_detection_performance}).
\begin{table}[H]
	\caption{Overall detection performance for initial contacts and final contacts as quantified by recall, precision and F$_{1}$ score.\label{tab:overall_detection_performance}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{r|CCCCCC|CCCCCC}
			\toprule
			 & \multicolumn{6}{c}{\textbf{Initial contacts}}	& \multicolumn{6}{c}{\textbf{Final contacts}}\\
			\textbf{Tracked point}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision}	& \textbf{F1}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision} 	& \textbf{F1}\\
			\midrule
			left ankle		& 624	& 19	& 5		& 97\%	& 99\%	& 98\%		& 606	& 32	& 10	& 95\%	& 98\%	& 97\%\\
			right ankle		& 599	& 42	& 8 	& 93\%	& 99\%	& 96\%		& 614	& 17	& 12	& 97\%	& 98\%	& 98\%\\
			left shank		& 605	& 38	& 15	& 94\%	& 98\%	& 96\%		& 585	& 53	& 18	& 92\%	& 97\%	& 94\%\\
			right shank		& 603	& 36	& 15 	& 94\%	& 98\%	& 96\%		& 595	& 30	& 9		& 95\%	& 99\%	& 97\%\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{TP: true positives, FN: false negatives, FP: false positives, F1: F$_{1}$ score.}}
\end{table}

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=18.5cm]{fig/bland_altman_plots_stride_params}
	\end{adjustwidth}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods.\label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

@@

Next, the time error, that is the difference between the annotated event and the detected event, was of interest. For both ICs and FCs, and for all sensor locations, the observed time error was small, and the middle $50\%$ of the time errors were within a range of $\left[-0.015,0.010\right]$ s (Table~\ref{tab:time_errors}, Figure~\ref{fig:time_error_box_plots}). These data showed that the deep learning-based approach is precise in detecting initial and final contacts. Time errors were slightly smaller than our previously reported results \cite{Romijnders2021} that used a heuristics-based approach \cite{Salarian2004}. The heuristics-based approach determined ICs and FCs as local minima in the medio-lateral angular velocity \cite{Salarian2004} and it could be that these minima do not exactly coincide with the referent event timing as determined from the OMC systems. The median and IQR of the time errors were in the same range as previously found by \cite{Gadaleta2019}. The time errors were lower than previously reported time errors from a continious wavelet-based approach \cite{Ji2019}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization, R.R., G.S., and W.M.; methodology, R.R., G.S.; software, R.R.; validation, R.R.,; formal analysis, R.R.; investigation, R.R., G.S., W.M.; resources, C.H, W.M..; data curation, E.W., R.R.,; writing---original draft preparation, R.R.; writing---review and editing, E.W., C.H., G.S., W.M.; visualization, R.R.; supervision, G.S., W.M.; project administration, R.R., E.W., C.H., G.S., W.M.; funding acquisition, W.M. All authors have read and agreed to the published version of the manuscript.}

\funding{This research received funding from the DFG Open Access fund.}

@@

\dataavailability{Data from the first 10 participant are available online at \url{https://github.com/neurogeriatricskiel/Validation-dataset}. Additionally, we are preparing the open-source release of all data from the healthy younger and older adults. Data from patient groups can be shared upon reasonable request.}

@@

	\bibitem[Mishra(2019)]{Mishra2019}
	Mishra,~P.; Pandey,~C.M.; Singh,~U.; Gupta,~A.; Sahu,~C.; Keshri,~A. Descriptive Statistics and Normality Tests for Statistical Data. {\em Ann. Card. Anaesth.} {\bf 2019}, {\em 22}(1), 67--72. 	
	\bibitem[Wilcoxon(1945)]{Wilcoxon1945}
	Wilcoxon,~F. Individual Comparisons by Ranking Methods. {\em Biometrics Bull.} {\bf 1945}, {\em 1}, 80--83.
	\bibitem[McDonald(2014)]{McDonald2014}
	McDonald,~J.H. \textit{Handbook of Biological Statistics}, 3rd ed.; Sparky House Publishing: Sparky House Publishing, USA, 2014; pp. 186--189.
\end{thebibliography}

>>>>

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{Many algorithms use 3D accelerometer and/or gyroscope data from inertial measurement unit (IMU) sensors to detect gait events (i.e., initial and final foot contact). However, these algorithms often require knowledge about sensor orientation and use empirically derived thresholds. As alignment cannot always be controlled for in ambulatory assessments, methods are needed that require little knowledge on sensor location and orientation, e.g., a convolutional neural network-based deep learning model. Therefore, 160 participants from healthy and neurologically diseased cohorts walked 5 meter distances at slow, preferred, and fast walking speed, while data were collected from IMUs on the left and right ankle and shank. Gait events were detected and stride parameters were extracted using a deep learning model and an optoelectronic motion capture (OMC) system for reference. Results showed a high detection rate for both initial and final contacts across sensor locations (recall $\ge 92\%$, precision $\ge 97\%$). Time agreement was excellent as witnessed from the median time error (0.005 s) and corresponding inter-quartile range (0.020 s). The extracted stride-specific parameters were in good agreement with parameters derived from the OMC system (maximum mean difference 0.003 s, and corresponding maximum limits of agreement (-0.049~s, 0.051~s) for a $95\%$ confidence level.) Thus, the deep learning approach was considered a valid approach for detecting gait events and extracting stride-specific parameters with little knowledge on exact IMU location and orientation in conditions with and without walking pathologies due to neurological diseases.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
Gait deficits are common in older adults and possibly reflect the presence of an underlying neurodegenerative disease \cite{Snijders2007,Hodgins2008}. For example, conversion to Parkinson's Disease \cite{DelDin2019} or conversion from mild cognitive impairment to Alzheimer's Disease \cite{Koenig2017,Bertoli2018} are linked with changes in spatiotemporal gait parameters. Similarly, temporal gait parameters are different for stroke patients \cite{SchroederVon1995,Mohan2021} and patients with multiple sclerosis \cite{Griskevicius2016,Flachenecker2019} when compared to healthy controls. To objectively quantify gait deficits, stride-specific parameters such as stride time or stride length are often used \cite{Hannink2016}. The beginning and end of a stride are determined from two successive initial contacts (ICs) of the same foot \cite{Perry2010,Whittle2012}. The IC is when the foot contacts the ground and together with the instant at which the foot leaves the ground (final contact, FC), each stride can be divided in a stance and swing phase \cite{Rueterbories2010,Bruening2014}.  The events of IC and FC, also referred to as \emph{gait events}, are commonly determined using force or pressure measuring devices \cite{Bruening2014}, or marker-based optoelectronic motion capture systems (OMC; henceforth referred to as the \emph{marker-based} system or method) \cite{Chiari2005,Topley2020}. These systems are relatively expensive, and restricted to usage in expertise laboratories \cite{Iosa2016,Jarchi2018}. As there is increasing evidence that gait measured in the lab does not reflect daily-life gait \cite{Hillel2019,Warmerdam2020,Atrsaei2021}, there is increasingly more interest in measurement systems that allow for continuous gait analysis in ambulatory settings. Therefore, the use of inertial measurement units (IMUs) is especially attractive, as these can be used to measure gait in ecologically valid environments, such as the home environment, thereby painting a more complete picture of health status \cite{DelDin2016,Shah2020} and providing clinical information that is complementary to standardized lab-based assessments \cite{Warmerdam2020,Fasano2020,Atrsaei2021,Corra2021}.

Previous research suggests that gait event detection is more accurate using an IMU worn on a lower limb (e.g., shank or foot) compared to an IMU worn on the low back \cite{BenMansour2015,Storm2016,Panebianco2018}. Now, in order to get from abstract IMU sensor readings to clinically relevant gait parameters (i.e., from accelerations and angular velocities to stride times) \cite{Hannink2016}, different algorithmic approaches have been developed in the last twenty years of clinical gait research. A recent study has evaluated a cross-section of these algorithms for different sensor locations on the lower leg and foot \cite{Niswander2021}. The algorithms were categorized according to which signals were analyzed, for example the angular velocity about the medio-lateral axis, or the accelerations along vertical and antero-posterior axes. This means the sensor readings need to be linked with the anatomical axes, that is, one needs to know which sensor axis aligns with for example the medio-lateral axis. In most approaches, it is simply assumed that due to sensor attachment the sensor axis aligns roughly with the anatomical axis of interest \cite{Salarian2004,Sabatini2005,Jasiewicz2006,Catalfamo2010,Trojaniello2014,Maqbool2016,Romijnders2021} or an additional calibration procedure (e.g., \cite{Ferraris1995}) is required \cite{Greene2010,Niswander2021}. In ambulatory assessments however, study participants often attach the sensor themselves, for example after showering, and therefore the sensor location and alignment cannot be controlled for. Furthermore, it is unlikely that each time the sensor is (re-)attached study participants, especially those with gait deficits, perform a calibration procedure that usually consists of holding a pre-defined pose and performing some known movement sequences \cite{Leineweber2019,Pacher2020}.

@@

For the current analysis data from four IMUs (Noraxon USA Inc., myoMOTION, Scottsdale, AZ, USA) were considered, namely those that were attached laterally above the left and right ankle joint and those attached proximally at the left and right shank. IMUs were secured to participants using elastic bands with a special hold for the IMU. Furthermore, reflective markers were attached on top of the usual foot wear at the heel and toe of both feet (Figure~\ref{fig:methods_marker_vs_imu}). Marker data were recorded using a twelve-camera OMC system (Qualisys AB, G\"{o}teborg, Sweden) at a sampling frequency of 200 Hz. IMU data were recorded at the same sampling frequency, and both systems were synchronized using a TTL signal \cite{Warmerdam2021}.

@@

All participant data data were split in three independent data sets, namely a training set, a validation set, and a test set. Each set contained data from approximately one-third of the participants. Participants were randomly assigned to one of the sets, stratified by both group (i.e., diagnosis) and gender (Table~\ref{tab:demographics_data}). The training and validation set were used to train an optimal deep learning model. Test set data were not used for training the model or hyperparameter tuning. The results on the model's performance were only based on the test set, and therefore reflected how good the model generalizes to new, \emph{unseen} data.

@@

\subsection{Model}
\subsubsection{Model Architecture}
\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=13.5cm]{fig/tcn_model_arch}
	\end{adjustwidth}
	\caption{The generic model architecture of the deep learning model to predict initial contacts (ICs) and final contacts (FCs). The inputs are the accelerometer and gyroscope data from a single inertial measurement unit, and are fed to a temporal convolutional network (TCN) (left). The TCN consisted of repeating residual blocks (ResBlocks) with exponentially increasing dilation factor (middle). Each ResBlock was built from two sequences of a convolutional layer (Conv), batch normalization layer (BatchNorm), a rectified linear unit activation layer (ReLU), and a dropout layer (DropOut) (right). \label{fig:tcn_model_arch}}
\end{figure}
The generic architecture for the deep learning model was based on a temporal convolutional network (TCN) \cite{Bai2018,Remy2020,Filtjens2020}. The TCN consisted of a sequence of residual blocks with exponentially increasing dilation factor \cite{YuKoltun2016,Bai2018}. Each residual block was built from two sequences of a dilated convolutional layer \cite{YuKoltun2016}, a batch normalization layer \cite{Ioffe2015}, a rectified linear unit (ReLU) activation layer, and a dropout layer \cite{Srivastava2014} (Figure). The model was built in Python \cite{VanRossum2009} using the high-level TensorFlow API Keras \cite{Remy2020,Chollet2015}.

@@

\subsection{Analysis}
The predictions of the model on the test set data were compared with the labels from the test set. The model performance was evaluated for (1) overall detection performance, (2) time agreement between the predicted events and the (marker-based) annotated events, (3) agreement between subsequently derived stride-specific gait parameters.
\subsubsection{Overall Detection Performance \label{subsubsec:overall_detection_performance}}
The overall detection performance quantified how many of the annotated evens were detected by the model (true positives, TP), how many of the annotated events were not detected (false negatives, FN), and how many events that were detected, were actually not annotated (false positives, FP). From these metrics, the recall (or sensitivity), precision and F$_{1}$ score were calculated as:
\begin{equation}
	\textrm{recall} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FN}} 
	\label{eqn:recall}
\end{equation}
\begin{equation}
	\textrm{precision} = \frac{\textrm{TP}}{\textrm{TP} + \textrm{FP}} 
	\label{eqn:precision}
\end{equation}
\begin{equation}
	\textrm{F$_{1}$ score} = 2 \cdot \frac{\textrm{recall} \cdot \textrm{precision}}{\textrm{recall} + \textrm{precision}} 
	\label{eqn:f1_score}
\end{equation}

\subsubsection{Time Agreement}
For all correctly detected gait events (TP, Section \ref{subsubsec:overall_detection_performance}), time agreement was assessed by the time error between the annotated and detect gait event, which was defined as:
\begin{equation}
	\textrm{time error} = t_{\textrm{ref}} - t_{\textrm{pred}}
	\label{eqn:time_agreement}
\end{equation}
with $t_{\mathrm{ref}}$ the gait event time from the marker-based annotations, and $t_{\mathrm{pred}}$ the gait event time from the model predictions. As a robust measure for the average time error and its spread, the median time error and the inter-quartile range (IQR) were reported \cite{OpenIntro2019}.

\subsubsection{Stride-specific Gait Parameters}
For those trials for which all gait events were detected, and no false positives were detected, the stride time, stance time and swing were calculated. Stride was the time between two successive ICs of the same foot. Stance time was the time between a FC and the preceding IC of the same foot. Swing time was the time between the IC following the last FC of the same foot.

@@

\subsection{Overall Detection Performance}
The performance of detecting ICs and FCs was objectively quantified by the number of TPs, the number of FNs, and the number of FPs. From these numbers, recall, precision and F$_{1}$ score were calculated (Table \ref{tab:overall_detection_performance}).
\begin{table}[H]
	\caption{Overall detection performance for initial contacts and final contacts as quantified by recall, precision and F$_{1}$ score.\label{tab:overall_detection_performance}}
	\begin{adjustwidth}{-\extralength}{0cm}
		\newcolumntype{C}{>{\centering\arraybackslash}X}
		\begin{tabularx}{\fulllength}{r|CCCCCC|CCCCCC}
			\toprule
			 & \multicolumn{6}{c}{\textbf{Initial contacts}}	& \multicolumn{6}{c}{\textbf{Final contacts}}\\
			\textbf{Tracked point}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision}	& \textbf{F1}	& \textbf{TP}	& \textbf{FN}	& \textbf{FP}	& \textbf{recall}	& \textbf{precision} 	& \textbf{F1}\\
			\midrule
			left ankle		& 624	& 19	& 5		& 97\%	& 99\%	& 98\%		& 606	& 32	& 10	& 95\%	& 98\%	& 97\%\\
			right ankle		& 599	& 42	& 8 	& 93\%	& 99\%	& 96\%		& 614	& 17	& 12	& 97\%	& 98\%	& 98\%\\
			left shank		& 605	& 38	& 15	& 94\%	& 98\%	& 96\%		& 585	& 53	& 18	& 92\%	& 97\%	& 94\%\\
			right shank		& 603	& 36	& 15 	& 94\%	& 98\%	& 96\%		& 595	& 30	& 9		& 95\%	& 99\%	& 97\%\\
			\bottomrule
		\end{tabularx}
	\end{adjustwidth}
	\noindent{\footnotesize{TP: true positives, FN: false negatives, FP: false positives, F1: F$_{1}$ score.}}
\end{table}

@@

\begin{figure}[H]
	\begin{adjustwidth}{-\extralength}{0cm}
		\centering
		\includegraphics[width=18.5cm]{fig/bland_altman_plots_stride_params}
	\end{adjustwidth}
	\caption{The agreement of extracted gait parameters between the sensor-based and marker-based methods. The differences between the stride-specific temporal gait parameters extracted from the marker-based and proposed approach are plotted against the means. \label{fig:gait_parameters_bland_altman_plots}}
\end{figure}

@@

Next, the time error, that is the difference between the annotated event and the detected event, was of interest. For both ICs and FCs, and for all sensor locations, the observed time error was small, and the middle $50\%$ of the time errors were within a range of $\left[-0.015,0.010\right]$ s (Table~\ref{tab:time_errors}, Figure~\ref{fig:time_error_box_plots}). These data showed that the deep learning-based approach is precise in detecting initial and final contacts. Time errors were slightly smaller than our previously reported results \cite{Romijnders2021} that used a heuristics-based approach \cite{Salarian2004}. The heuristics-based approach determined ICs and FCs as local minima in the medio-lateral angular velocity \cite{Salarian2004} and it could be that these minima do not exactly coincide with the referent event timing as determined from the OMC systems. The median and IQR of the time errors were in the same range as previously found by \cite{Gadaleta2019}. The time errors were lower than previously reported time errors from a continuous wavelet-based approach \cite{Ji2019}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization, R.R., G.S., and W.M.; methodology, R.R., G.S.; software, R.R.; validation, R.R.; formal analysis, R.R.; investigation, R.R., G.S., W.M.; resources, C.H, W.M.; data curation, E.W., R.R.,; writing---original draft preparation, R.R.; writing---review and editing, E.W., C.H., G.S., W.M.; visualization, R.R.; supervision, G.S., W.M.; project administration, R.R., E.W., C.H., G.S., W.M.; funding acquisition, W.M. All authors have read and agreed to the published version of the manuscript.}

\funding{}

@@

\dataavailability{Data from the first 10 participant are available online at \url{https://github.com/neurogeriatricskiel/Validation-dataset}. Additionally, we are preparing the open-source release of all data from the healthy younger and older adults. Data from patient groups can be shared upon reasonable request. The scripts are publicly available at the author's personal GitHub, which can be found at \url{https://github.com/rmndrs89/my-gait-events-tcn}.}

@@

%	\bibitem[Mishra(2019)]{Mishra2019}
%	Mishra,~P.; Pandey,~C.M.; Singh,~U.; Gupta,~A.; Sahu,~C.; Keshri,~A. Descriptive Statistics and Normality Tests for Statistical Data. {\em Ann. Card. Anaesth.} {\bf 2019}, {\em 22}(1), 67--72. 	
%	\bibitem[Wilcoxon(1945)]{Wilcoxon1945}
%	Wilcoxon,~F. Individual Comparisons by Ranking Methods. {\em Biometrics Bull.} {\bf 1945}, {\em 1}, 80--83.
%	\bibitem[McDonald(2014)]{McDonald2014}
%	McDonald,~J.H. \textit{Handbook of Biological Statistics}, 3rd ed.; Sparky House Publishing: Sparky House Publishing, USA, 2014; pp. 186--189.
\end{thebibliography}
